---
format: 
  html:
    toc: false
    theme: yeti
---
## Key for workshop 5 exercises
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, cache = T)
```

1)  Using the sparrows dataset from the primer package, plot the population size over time. Calculate the average $\lambda$ (hint: you will need to use the geometric mean) for the sparrow population.
```{r}
library(tidyverse)
library(primer)
library(DescTools)
data('sparrows')

ggplot(data=sparrows, aes(x=Year, y=Count)) +
  geom_line() + 
  geom_point() +
  theme_bw()

df = sparrows |>  
  mutate(lambda = lead(Count)/Count) |> 
  summarize(mean = Gmean(lambda, na.rm = T),
            sd = Gsd(lambda, na.rm = T))

```


2)  Project the population over the next 50 years for the sparrow population using the average $\lambda$ from exercise 1. For now use the mean sparrow count for $K$, and use the minimum sparrow count as $N_0$. Plot the results.
```{r}
lambda = df$mean
rd = lambda - 1
K = mean(sparrows$Count)
N0 = min(sparrows$Count)
years = 50
  
pop = tibble(time = 0:years, Nt = NA) 
pop$Nt[pop$time == 0] = N0

# for loop
for(i in 2:length(pop$time)){
  pop$Nt[i] = pop$Nt[i - 1]  + rd*pop$Nt[i - 1]*(1 - pop$Nt[i - 1] / K)
}

ggplot(data=pop, aes(x=time, y=Nt)) +
  geom_line() + 
  geom_point() +
  theme_bw()
```


3)  Did it reach $K$? If not, how many years does it take based on your $r_d$? Plot the results.
```{r}
lambda = df$mean
rd = lambda - 1
K = mean(sparrows$Count)
N0 = min(sparrows$Count)
years = 10000

pop = tibble(time = 0:years, Nt = NA) 
pop$Nt[pop$time == 0] = N0

# for loop
for(i in 2:length(pop$time)){
  pop$Nt[i] = pop$Nt[i - 1]  + rd*pop$Nt[i - 1]*(1 - pop$Nt[i - 1] / K)
}

ggplot(data=pop, aes(x=time, y=Nt)) +
  geom_line() + 
  geom_point() +
  theme_bw()
```


4)  Challenge: Simulate/Project the sparrow based on different $K$ = 25, 30, 40, 50. Plot the results.
```{r}
lambda = df$mean
rd = lambda - 1
K = c(25,30,40,50)
N0 = min(sparrows$Count)
years = 500

pop = tibble(time = rep(0:years, times = length(K)), Nt = NA, K = rep(K, each = years+1)) 
pop$Nt[pop$time == 0] = N0

# for loop
for(i in 1:nrow(pop)){
  if(pop$time[i] > 0){
    pop$Nt[i] = pop$Nt[i - 1]  + rd*pop$Nt[i - 1]*(1 - pop$Nt[i - 1] / pop$K[i])
  }
}

ggplot(data=pop, aes(x=time, y=Nt, color = as.factor(K))) +
  geom_line() + 
  geom_point() +
  theme_bw()

```
