---
format: 
  html:
    toc: false
    theme: yeti
---
## Key for workshop 1 exercises
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, cache = T)
```


1. Make two vectors, object `a` containing the values 2, 3, 4, and 5 and object `b`containing the values 50, 100, 38, and 42.
```{r}
a = c(2:5)
a

b = c(50, 100, 38, 42)
b
```

2. Multiply object `a` by 3 and assign it to a new object, divide object `b` by 5 and assign it to a new object, then add the new two objects together. 
```{r}
c = a*3
c

d = b/5
d
```


3. Create a new `data.frame`/`tibble` with the four objects created above

```{r}
#base R
df_ex1 = data.frame(a = a, b = b, c = c, d = d)
df_ex1

library(tidyverse)
#tidyverse
df_ex1 = tibble(a = a, b = b, c = c, d = d)
df_ex1
```


4. Save the `data.frame`/`tibble` created in exercise 3 as a .csv
```{r, eval = F}
#base R
write.csv(df_ex1, "./data/df_ex1.csv") #path name will be  case specifice - i.e., depending on the coder's working directory

# tidyverse
write_csv(df_ex1, 'data/df_ex1.csv')
```

5. Load in files a.csv and b.csv (found on [github](https://github.com/BSC6926-B52-Fall2023/workshopScripts/tree/main/data) and canvas) and assign each as an object.
```{r, eval = F}
#base R
a_new = read.csv("data/a.csv")
b_new = read.csv("data/b.csv")

# tidyverse
a_new = read_csv("data/a.csv")
b_new = read_csv("data/b.csv")
```


