---
format: 
  html:
    toc: false
    theme: yeti
---
## Key for workshop 3 exercises
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, cache = T)
```

1. Using the iris data create a new tibble or dataframe that calculates the median value of all petal and sepal measurements for each species. 
```{r}
library(tidyverse)

df = iris |> 
  group_by(Species) |> 
  summarize(across(where(is.numeric), median))

df = iris |> 
  group_by(Species) |> 
  summarize(across(Sepal.Length:Petal.Length, median))

```


2. Simulate population growth over 25 years of a population that has random growth from a mean of 10 individuals and a sd of 5 individuals that starts at a population size of 100. 
```{r}
start = 100 
steps = 25
pop = tibble(time = 0:steps, n = NA) #Output vector size 10
pop$n[pop$time == 0] = start

for (t in 1:steps){ #sequence 
  growth = rnorm(n =1, mean = 10, sd = 5) #extracting random value normal dist
  pop$n[pop$time == t] = growth + pop$n[pop$time == (t-1)]
}
```


3. Plot population size over time
```{r}
ggplot(pop, aes(time, n))+
  geom_point(size = 2)+
  geom_line(linewidth = 1)+
  labs(x = 'Time', y = expression(italic(N[t])))+
  theme_classic()+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))

```


4. Create a quarto or rmarkdown document of the exercises.